---
title: "Reproducible Research Assignment 1"
author: "Kevin Creese"
date: "Wednesday, September 16, 2015"
output: html_document
---

## Summary
This report is an R Markdown file that describes the process used to compile data presented in the first assignment of **Coursera's Reproducible Research**.  Throughout this assignment we will identify the steps used to:
        - Load and process the data
        - Answer the question: What is the mean total number of steps taken per day?
        - Answer the question: What is the average daily activity pattern?

### Data Overview
The data used for this report comes from the activities recorded by an unnamed individual wearing a personal movement device such as a [Fitbit](http://www.fitbit.com), [Nike Fuleband](http://www.nike.com/us/en_us/c/nikeplus-fuel), or [Jawbone Up](https://jawbone.com/up). This assignment makes use of data from a personal activty monitoring device.  The device collects data in five minute intervals through out the day.  The data consists of two months of data from an anonymous individual collected during teh months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.  

### Data Collection
The data used in this report can be can be accessed from [Coursera's Reproducible Research - Assignment 1](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

The variables included in this data set are: 
<tr>
   1. **steps:**Number of steps taking in a 5-minute interval (missing values are recorded as NA)  
   2. **date:** The date on which the measurement was taken in YYYY-MM-DD format  
   3. **interval:** Identifier for the 5-minute interval in which measurements was taken  
   
The dataset is stored in a comma-seperated-value (CSV) file and there are a total of 17,568 observations contained within the dataset.

### Load Data - Download, set dataframe, view summary of data
*First we will setup a tempfile and download and unzip the dataset.  Once that is done we will assign the data set to the dataframe `data`.*
````{r LoadDataSet, echo=TRUE}
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
        temp <- tempfile()
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
        unzip(temp)
        unlink(temp)
}

data <- read.csv("activity.csv")
summary(data)
```


## What is the mean total number of steps taken? 
*(raw data)*
Summarize the steps by day, create a histogram, and calculate the mean and median.
```{r RawMean-Hist}
dailySteps <- aggregate(steps ~ date, data, sum)
hist(dailySteps$steps, main = paste("Total Steps by Day"), xlab="Number of                 Steps Each Day", col="green")
StepsMean <- mean(dailySteps$steps)
StepsMedian <- median(dailySteps$steps)
```

The mean is `r StepsMean` and the median is `r StepsMedian`

## What is the average daily activity pattern?

 - Calculate the average steps for each interval for all days.  
 - Plot the average number of steps per day by interval.  
 - Find intervals with the most average steps.  

```{r intervalsteps}
intervalSteps <- aggregate(steps ~ interval, data, mean)

plot(intervalSteps$interval,intervalSteps$steps, type="l", xlab="Interval", ylab="Number of Steps", main="Ave Number of Steps per Day by Interval")

intervalMax <- intervalSteps[which.max(intervalSteps$steps),1]
```

## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

The average 5-minute interval that contains the maximum number of steps is `r intervalMax`.

## Imputing Missing Values
*There are a number of days and intervals that are missing values.  These have been captured as 'NA'. This section of the report will:*  
 -- Capture and report the total number of missing values in the datase by reporting the total number of rows with 'NA's.
 -- Devise a strategy for filling in all of these missing values.  Don't get to fancy, so I'm going to use the KISS method for this and apply the average for each day to a missing interval.
 -- Then create a new data set that is equal to the origonal data set, but with the missing values built in.  
 -- Make a histogram of the total number of steps taken each dan and calculate and report the **mean** and **median** total number of steps taken per day.    
 - Do these values differ from the estimates from the first part of the assignment?  
 - What is the impact of imputting missing data on the estimates of the total daily number of steps?  

### Total number of missing values
```{r missingValues}
missing <- sum(!complete.cases(data))           
```
The total number of missing values is `r missing`.

The stategy used to insert missing data will be to input and average for each interval.
```{r modData}
modData <- transform(data, steps = ifelse(is.na(data$steps), intervalSteps$steps[match(data$interval, intervalSteps$interval)], data$steps))
summary(modData)

```

Recalculate the daily steps with the revisions made.

```{r modDailySteps}
modDailySteps <- aggregate(steps ~ date, modData, sum)
hist(modDailySteps$steps, main = paste("Modified - Total Steps Each Day"), col="red", xlab="Number of steps")
```

Calculate the mean and median with the revised data
```{r}
rmain <- mean(modDailySteps$steps)
rmedian <- median(modDailySteps$steps)
```

So...  What's the diff between regular and revised?
```{r}
mean_diff <- rmain - StepsMean
median_diff <- rmedian - StepsMedian
```

So...  What's the total difference?
```{r}
total_diff <- sum(modDailySteps$steps) - sum(dailySteps$steps)
```

The revised data mean is `r rmain`.  
The revised data median is `r rmedian`.  
The difference between the regular mean and the revised mean is `r mean_diff`.  
The difference between the regular median and the revised median is `r median_diff`.  
The difference between the total number of steps between regular and unleaded data is `r total_diff`.   

The result of modifying the data was that there were `r total_diff` more steps in the revised data.


### Are there difference in activity patterns between weekdays and weekends?
* The last part of this assignment is to evaluate the number of steps over the weekend verses weekdays.  We are going to create two factor variables called weekday and weekend.  By evaluating each day to determine if it is a weekend or weekday we can then make a plot to compair the the average number of steps between each data set.*

Create the new factor variables
```{r}
weekdays <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
modData$dow <- as.factor(ifelse(is.element(weekdays(as.Date(modData$date)),weekdays), "Weekday", "Weekend"))
dailySteps_2 <- aggregate(steps ~ interval + dow, modData, mean)

library(lattice)
xyplot(dailySteps_2$steps ~ dailySteps_2$interval|dailySteps_2$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")


```

The End...